Mergeable Types

### 2017-02-28 paper/appendix.tex
Committing meta.

\section{Appendix}
\input{list-code}
\begin{lemma} [\bfseries Unique External Locus]
Every version $v$ (except the \C{INIT} version) in the history $H$
generated by the rules in Fig.~\ref{fig:opsem} has a unique external
locus.
\end{lemma}
\begin{proof}
By case analysis on the reduction relation. In each case, we assume
that the unique external locus property holds for the initial history,
and prove that it holds for the final history.

In \rulelabel{E-Run} case, initial version is the unique external locus of
the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
which creates the branch $H(t')$ by forking from $H(t)$, the latest
version on $H(t)$ is the external locus of the first version on
$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
external locus as the previous latest version, which, premise says, is
unique. No new version is created in \rulelabel{E-Pull}. In
\rulelabel{E-Pull-Wait}, new version is created by merging the latest
version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
Premise says that $v$ has a unique external locus (call it $v_o$). The
new version $v_m$ inherits $v_o$ as a least external ancestor, and
also has $v'$ as another least external ancestor. However, since
$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
have $v'$ as the the external locus.
\qed
\end{proof}

\begin{lemma} [\bfseries Unique LCA]
Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
the rules in Fig.~\ref{fig:opsem} have a unique least
common ancestor. 
\end{lemma}
\begin{proof}
By case analysis on the step relation. In each case, we assume that
the unique LCA property holds for the initial history, and prove that
it holds for the final history.

The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
Let $v_1$ be the previous version on the branch, and let $v_2$ be any
other version. The premise is that $v_1$ and $v_2$ have a unique LCA
$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
locus is same as that of $v_1$. Hence, all ancestors of $v_1$
(including itself) are also ancestors of $v$. Moreover, $v$ has no new
ancestors except itself, which is the latest version being added.
Hence, any common ancestor of $v_1$ and $v_2$ is also a common
ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
Hence LCA of $v$ and $v_2$ is unique. The reasoning for
\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.

The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
$v'$ is merging into version $v$ to produce the next version $v_m$.
Let $v_2$ be any version. From the premise, we know that $v$ and $v_2$
have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
LCA (call it $v_1$). But, we know that $v_2$ has a unique external
locus (call it $v_3$), hence we know that 
\begin{itemize}
  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
  $v_2$ and $v_m$ have a unique LCA.

  \item $v_0$ is an internal ancestor of $v_2$, hence an external
  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
\end{itemize}
\end{proof}

>>>>

\section{Appendix}
\input{list-code}
\begin{lemma} [\bfseries Unique External Locus]
Every version $v$ (except the \C{INIT} version) in the history $H$
generated by the rules in Fig.~\ref{fig:opsem} has a unique external
locus.
\end{lemma}
\begin{proof}
Follows from the definitions~\ref{def:external-locus},
\ref{mergeability}, and the \rulelabel{E-Pull-Wait} rule.
\end{proof}

\begin{lemma} [\bfseries Unique LCA]
Every pair of branches in a legal history $H$, if mergeable as per
Def.~\ref{fig:mergeability}, have a unique least common ancestor. 
\end{lemma}
\begin{proof}
By construction. If $b_j$ is mergeable into $b_i$, the locus of
$b_i$ is the least common ancestor of $b_i$ and $b_j$
\end{proof}

\begin{lemma}
Every pair of 
\end{lemma}
% If $b_j$ is not mergeble into $b_i$, it could be because:
% 1. $b_j$ included some internal nodes of $b_i$ that are not in
% $L(b_i)$, then simply merge $b(L(b_i))$ into $b_i$, and then that back
% into $b(L(b_i))$. Now the problem reduces to merging $b(L(b_i))$
% 2. $If

### 2017-02-28 paper/meta.tex
Committing meta.

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_3$ at the instance of
merging $b_1$. Since it has already merged $b_2$, a version on $b_2$
($v_{21}$) could be a common ancestor for $b_3$ and some other branch
(call it $b$). Now, if $b_3$ merges $b_1$, same could be true of $b_3$
and $b_1$: a version on $b_1$ ($v_{11}$) could be a common ancestor
for $b_3$ and the other branch $v$. Since $v_{21}$ is also a common
ancestor for both the branches, and the ancestors are not ordered by
the ancestor relation, $b_3$ and $b$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $b$ is played both by $b_1$
and $b_2$. We observe that this scenario can be prevented if, when
merging $b_1$, $b_3$ insists on an ancestor relation between the last
merged version ($v_{22}$ of the $b_2$), and the latest version of
$b_1$, the currently merging branch. We call the last merged version
the \emph{external locus} of the branch. By requiring that, for every
branch $b$, the external locii of $b$ at various points in time (i.e.,
external locus of every prefix of $b$) be totally ordered, and given
that the semantics already preempts criss-cross merges, we effectively
avoid the case of there ever being more than 2 LCAs for a pair of
branches. Thus, uniqueness of LCAs is thus guaranteed.

We now formalize the notion of mergeability, and the guarantees
provided by the operational semantics.

@@

\begin{definition} [\bfseries Locus]
Given a branch $b$ and a version $v\in b$, locus ($v_o$) of $v$ is an
ancestor that is not an ancestor of any other external ancestor of
$v$. That is, $\under{H}{v_o \preceq v}$, and there does not exist a
$v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
$\under{H}{v_o \preceq v_o'}$. We write $L(v)$ to denote the locus
of version $v$.
\end{definition}

In a legal branching history, every version has a unique
locus\footnote{lemma and proof in the appendix}. Thus, locus of a
version is the maxima of its causal history. As usual, the locus of
the latest version on a branch is called the locus of the branch.

\begin{definition} [\bfseries Mergeability]
\label{def:mergeability}
Given a history $H$, a version $v_1$ and a version $v_2$ that is not
an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
$\under{H}{v_2 \mbleto v_1}$) if and only if $L(v_1)$ is an ancestor
of $v_2$, and no internal ancestor of $v_1$ that is not also an
ancestor of $v_o$ is an ancestor of $v_2$. 
If $v_1$ and $v_2$ are the latest versions on their respective
branches $b_1$ and $b_2$, we say $b_2$ is mergeable into $b_1$, or
more generally, $b_1$ and $b_2$ are mergeable.  
\end{definition}

The following theorem asserts that every pair of branches deemed
mergeable by the above definition are infact mergeable, i.e., they are
guaranteed to have a unique LCA.

\begin{lemma} [\bfseries Unique LCA]
Every pair of branches in a legal history $H$, if mergeable as per
Def.~\ref{fig:mergeability}, have a unique least common ancestor.
\end{lemma}

\begin{lemma} [\bfseries Non-Monotonicity of Mergeability]
In a legal branching history $H$ produced by the operational
semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
Def.~\ref{def:mergeability}), then there exist a series of merges
involving only the mergeable branches that can be performed on $H$ to
yield a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
\end{lemma}

Thus, a pair of branches that are not currently mergeable (as per
Def.~\ref{fig:mergeability}), will become mergeable as the branching
history is extended.

\begin{theorem} [\bfseries Progress and Eventual Convergence] Given a
legal branching history $H$, and a program $p$, where every thread
expression in $p$ is a \C{pull}, there exist a series of reduction steps
that reduce $p; H$ to $p'; H'$, where every thread expression in $p$
is the same value $v$.
\end{theorem}

>>>>

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_3$ at the instance of
merging $b_1$. Since it has already merged $b_2$, a version on $b_2$
($v_{21}$) could be a common ancestor for $b_3$ and some other branch
(call it $b$). Now, if $b_3$ merges $b_1$, same could be true of $b_3$
and $b_1$: a version on $b_1$ ($v_{11}$) could be a common ancestor
for $b_3$ and the other branch $v$. Since $v_{21}$ is also a common
ancestor for both the branches, and the ancestors are not ordered by
the ancestor relation, $b_3$ and $b$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $b$ is played both by $b_1$
and $b_2$. We observe that this scenario can be prevented if, when
merging $b_1$, $b_3$ insists on an ancestor relation between the last
merged version ($v_{22}$) and the currently merging version.  We call
the last merged version of a branch its \emph{locus}. By maintaining a
total order of among locii of a branch $b$ as new versions are
created, we effectively guarantee the presence of a version (the
locus) that is lower than all the ancestors of the latest version on
$b$. Next, by requiring the locus to be an ancestor of the merging
version, we ensure that all the common ancestors have a single lowest
element, thus enforcing the uniqueness of the LCA.

We now formalize the intuitions described above via a series of
definitions that help us state the guarantees offered by the system.

@@

\begin{definition} [\bfseries Locus]
Given a branch $b$ and a version $v\in b$, locus ($v_o$) of $v$ is an
ancestor that is not an ancestor of any other external ancestor of
$v$. That is, $\under{H}{v_o \preceq v}$, and there does not exist a
$v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
$\under{H}{v_o \preceq v_o'}$. We lift the notion of locus to the
level of branches by defining the locus of a branch as the locus of
its latest version.
\end{definition}

In a legal branching history, every version has a unique locus
(property follows from Def.~\ref{def:mergeability} below). Let
\emph{causal history} of a version be defined as the set of all
ancestors of that version. A locus ($v_o$) of a version ($v$) is
therefore the maximum element (lowest element in the ancestor
relation) of the causal history of $v$'s internal ancestor that was
the result of the last merge. We make use of this observation while
defining mergeability.

\begin{definition} [\bfseries Mergeability]
\label{def:mergeability}
Given a history $H$, a version $v_1$ and a version $v_2$ that is not
an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
$\under{H}{v_2 \mbleto v_1}$) if and only if the locus ($v_o$) of
$v_1$ is an ancestor of $v_2$, and no internal ancestor of $v_1$ that
is not also an ancestor of $v_o$ is an ancestor of $v_2$.    
\end{definition}

If $v_1$ and $v_2$ referred in the above definition are the latest
versions on their respective branches $b_1$ and $b_2$, we say $b_2$ is
mergeable into $b_1$, or more generally, $b_1$ and $b_2$ are
mergeable. The definition essentially requires all of $v_1$'s causal
history that preceeds its locus $v_o$ to be included in $v_2$'s causal
history, and none of $v_1$'s history that succeeds $v_o$ to be
included in $v_2$'s history. This allows us to view versions $v_1$ and
$v_2$ as having been independently evolved from a common causal
history whose maximum (w.r.t the ancestor relation) is the version
$v_o$. Consquently, $v_o$ becomes the LCA for $v_2$'s merge into
$v_1$. Generalizing this observation for the latest versions on
any two branches, we state the following theorem:

\begin{theorem} [\bfseries Unique LCA]
Every pair of branches in a legal history $H$, if mergeable as per
Def.~\ref{fig:mergeability}, have a unique least common ancestor.
\end{theorem}

While the definition of mergeability is sufficient to enforce
uniqueness of LCAs, it may hinder progress in the sense that there may
not exist a pair of branches in a legal history satisfying the
definition, thus making the progress impossible. The following theorem
asserts that this is not the case:

\begin{theorem} [\bfseries Progress]
In a legal branching history $H$ produced by the operational
semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
Def.~\ref{def:mergeability}), then there exist a sequence of fork and
merge operations (between mergeable versions) that can be performed on
$H$ to yeild a new history $H'$, where $b_i$ and $b_j$ are mergeable.  
\end{theorem}

Proof is by the induction on the size of branching history. The
observation that the smallest history that is the union of causal
histories of two unmergeable versions (call them $v_i$ and $v_j$) is
smaller than the history that includes the unmergeable versions. Since
causal histories have maximum versions (respective locii), which, by
inductive hypothesis, can be made mergeable, we merge them to yeild a
new version $v$ that is the maximum of the causal histories of both
$v_i$ and $v_j$. If $v_i$ and $v_j$ are the latest versions on their
respective branches, the branches are now mergeable.

% Observe that $v$ and $v_i$ are mergeable (with
% $v_i$'s locus as the LCA), and so do $v$ and $v_j$. If $v_i$ and $v_j$
% are the latest versions on their respective branches, then merging $v$
% into $v_i$ and $v$ into $v_j$ makes $v$ the locus of both the
% branches, thus making them mergeable again.

% \begin{theorem} [\bfseries Progress and Eventual Convergence] Given a
% legal branching history $H$, and a program $p$, where every thread
% expression in $p$ is a \C{pull}, there exist a series of reduction steps
% that reduce $p; H$ to $p'; H'$, where every thread expression in $p$
% is the same value $v$.
% \end{theorem}
